Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-16 paper/main.tex
update todos

%###################################################################################################
\section{Introduction}
\textbf{
    Mein Vorgehen: ich passe Formulierungen an, wenn ich im Prinzip mit der Aussage einverstanden
    bin, aber den Wortlaut verbesserungswürdig finde.
    Ich schreibe Kommentare, dort wo ich eher grundsätzliche Schwierigkeiten sehe.
    Dort wo ich nichts geändert habe und keinen Kommentar gegeben habe, kann ich zu einem späteren
    Zeitpunkt doch Änderungen vornehmen wollen.
}

\acp{abm} have proven to be powerfull tools in the study of complex cellular
systems~\cite{Pleyer2023}.
They have been applied in various scenarios, such as cancer
research~\cite{Ghaffarizadeh2018,Cooper2020}, Immuno-Oncology~\cite{Karolak2021} or the modeling
of mechanical intercellular connections in plant cells \textbf{TODO CITATION}.
By treating cells as individual entities, these tools allow researchers to think in terms of
cellular properties and processes thus allowing for a better understanding of the biological
reality.

@@

@@

@@

Since this model is void of any cell-cell interactions or cell division, we are left with 5 parameters:
Magnitude of the drag force $C$, growth rate $\alpha$, rigidity of the rod $\eta$, spring tension
$\gamma$ and damping constant $\lambda$ (see table~\ref{table:list-of-parameters}).
Figure~\ref{fig:amir-bending-simulation} (B) shows a snapshot of the initial (orange) and final
(blue) position of the rod using the optimized parameters.
While the parameters for the dragging force, growth rate and damping can be identified, the same
can not be said for the rod rigidity and spring tension, which both are not identifiable.
However this behaviour is expected since, the shape of the rod is almost exclusively straight, thus
adding no further information except that a lower threshold of rigidity and spring tension should be
present.
\textbf{[Ist mir nicht ganz klar]}

Note that each point represented in the subfigures~\ref{fig:amir-bending-simulation} (C-G) was
optimized again, meaning that higher values of rigidity and spring tension can also lead to higher
values of drag forces.
\textbf{[Ist mir nicht ganz klar.]}

@@

To remove this non-identifiability we reduce the number of parameters by setting $\gamma=\gamma_0/s$
and $\eta=\eta_0 s$ for the rod rigidity and spring tension, resp.
We then optimize for this parameter and pick $\gamma_0=\SI{15}{\micro\metre}$ with
$\eta_0=\SI{150}{\micro\metre\per\second\squared}$ which corresponds to $60\%$ of the previously considered
interval.
The corresponding plots can be seen in \ref{subsec:supplement-parameter-profiles-amir}.

\begin{figure}[H]
    \centering
    \begin{minipage}{0.38\textwidth}
        \begin{tikzonimage}[width=\textwidth]
            {figures/amir-elastic-000032-wider.png}%
            \node at (0.01, 0.01+17pt)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
        \end{tikzonimage}%
    \end{minipage}%
    \begin{minipage}{0.62\textwidth}
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/fit-comparison.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/drag force.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/growth rate.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
        \end{tikzonimage}\\
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/rod rigidity.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/spring tension.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/damping.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{G}};
        \end{tikzonimage}%
    \end{minipage}
    \caption{
        \textbf{TODO remove pixels in units}
        \textbf{@Christian is it okay to change the units here to seconds from minutes? (compare
        table above)}
        Subfigure (A) shows a snapshot of the experiments performed by \cite{Amir2014_2}.
        Division of the bacteria is inhibited and they are partially trapped within a mother machine
        such that a controlled flow can be exerted on them.
        (B) A comparison of the initial and bent state show good visual agreement of our prediction
        with the experimental data.
        Subfigures (C-G) show likelihood profiles for the individual parameters.
    }
    \label{fig:amir-bending-simulation}
\end{figure}

@@

@@

We have already suspected (see Table~\ref{table:simulation-aspects}) that in order to properly
estimate parameters of our system, we need to account for natural cell-variability, thus assigning
parameters on a per-cell basis.
We have previously presented a list of $N_p$ parameters (see Table~\ref{table:list-of-parameters})
which will either be assigned individually or globally.
We filter out a subset $N_g\leq N_p$ of global parameters, treating the remaining ones individually.
Given $N_c$ cells, the total amount of parameters to be estimated is:
\begin{equation}
    N_{p,\text{tot}} = N_g + N_c (N_p - N_g).
\end{equation}
It is clear, that we should avoid, to needlessly treat parameters non-globally since this increases
the size of the parameter space considerably.
The volume that needs to be sampled to optimize within a given parameter space scales exponentially
with the dimension ie. the total number of parameters $N_{p,\text{tot}}$.
\textbf{[Das ist leider komplett unverständlich. Ich glaube, ich weiss schon, was du sagen möchtest, aber das muss umformuliert werden. Stichworte sind hier: Cell Variability, Heterogeneity, latent variables. Parameter per cell and individual parameters versteht auch niemand den Unterschied.]}
\textit{Ich habe versucht, es klarer zu machen. Bitte dazu Feedback!}

@@

@@

% --------------------------------------------------------------------------------------------------
\subsection{Without Division}
In this section we will focus on a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant direct comparison scheme (see
\ref{paragraph:direct-comparison}) for our cost function.
We have chosen said interval such that effects related to the lag phase can be neglected which
allows us to sample a lower amount of paremeters thus improving runtime performance of our
optimization scheme.

>>>>

\begin{itemize}
    \item Adapt introduction; \acp{abm} not focus of paper, but rather tool to do the job
    \begin{itemize}
        \item model division, growth of bacteria
        \item \acp{abm} are good for this tool
    \end{itemize}
    \item Exchange $\sigma \rightarrow R_1+R_2$
    \item Equation 8 make double norm $||\vec{F}||$
    \item Remove auxiliary variable $\vec{F}$ only use $\vec{\nabla}V$
    \item Table 2 in supplement? (probably not)
    \item numerical implementation verschieben?
    \item numerical implementation: waht do we need to simulate this?
    \begin{itemize}
        \item \acp{abm} are common tools to realize this
        \item cellular\_raza is THE tool to do it
    \end{itemize}
    \item Methodology: single-cell parameter estimation as motivation for techniques
    \item Which parameters are global which are individual? Discuss this in general; not in detail
    \item depends on data, number of parameters (computational efficiency)
    \item page 12; yes this is Discussion
    \item page 12: maybe extend the paragraph wrt comparison in intervals
\end{itemize}

%###################################################################################################
\section{Introduction}
\acp{abm} have proven to be powerfull tools in the study of complex cellular
systems~\cite{Pleyer2023}.
They have been applied in various scenarios, such as cancer
research~\cite{Ghaffarizadeh2018,Cooper2020}, Immuno-Oncology~\cite{Karolak2021} or the modeling
of mechanical intercellular connections in plant cells \textbf{TODO CITATION}.
By treating cells as individual entities, these tools allow researchers to think in terms of
cellular properties and processes thus allowing for a better understanding of the biological
reality.

@@

\textbf{TODO adapt naming to appliations section}

@@

\begin{itemize}
    \item go from simple to more complex: single-cell, many cells, with division
    \item explain data in a generalized way; detailed explanations within subsections
\end{itemize}

@@

Since this model is void of any cell-cell interactions or cell division, we are left with 5 parameters:
Magnitude of the drag force $C$, growth rate $\alpha$, rigidity of the rod $\eta$, spring tension
$\gamma$ and damping constant $\lambda$ (see table~\ref{table:list-of-parameters}).
Figure~\ref{fig:amir-bending-simulation} (B) shows a snapshot of the initial (orange) and final
(blue) position of the rod using the optimized parameters.
While the parameters for the dragging force, growth rate and damping can be identified, the same
can not be said for the rod rigidity and spring tension, which both are not identifiable.
However this behaviour is expected since, the shape of the rod is almost exclusively straight, thus
adding no further information except that a lower threshold of rigidity and spring tension should be
present.

Note that each point represented in the subfigures~\ref{fig:amir-bending-simulation} (C-G) was
optimized again, meaning that higher values of rigidity and spring tension can also lead to higher
values of drag forces.

@@

\textbf{TODO what is the parameter $s$? Explain more clearly.}\\
To remove this non-identifiability we reduce the number of parameters by setting $\gamma=\gamma_0/s$
and $\eta=\eta_0 s$ for the rod rigidity and spring tension, resp.
We then optimize for this parameter and pick $\gamma_0=\SI{15}{\micro\metre}$ with
$\eta_0=\SI{150}{\micro\metre\per\second\squared}$ which corresponds to $60\%$ of the previously considered
interval.
The corresponding plots can be seen in \ref{subsec:supplement-parameter-profiles-amir}.
\textbf{TODO Make this more clear. Transient information would lead to more information. We only
have initial and final snapshots.}

\begin{figure}[H]
    \centering
    \begin{minipage}{0.38\textwidth}
        \begin{tikzonimage}[width=\textwidth]
            {figures/amir-elastic-000032-wider.png}%
            \node at (0.01, 0.01+17pt)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
        \end{tikzonimage}%
    \end{minipage}%
    \begin{minipage}{0.62\textwidth}
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/fit-comparison.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/drag force.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/growth rate.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
        \end{tikzonimage}\\
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/rod rigidity.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/spring tension.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/damping.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{G}};
        \end{tikzonimage}%
    \end{minipage}
    \caption{
        \textbf{TODO rework caption}
        Subfigure (A) shows a snapshot of the experiments performed by \cite{Amir2014_2}.
        Division of the bacteria is inhibited and they are partially trapped within a mother machine
        such that a controlled flow can be exerted on them.
        (B) A comparison of the initial and bent state show good visual agreement of our prediction
        with the experimental data.
        Subfigures (C-G) show likelihood profiles for the individual parameters.
    }
    \label{fig:amir-bending-simulation}
\end{figure}

@@

\textbf{Possibly explain data more clearly}

\begin{itemize}
    \item show dataset (few snapshots)
\end{itemize}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/crm_fit-data-progression.png}
    \caption{TODO}
\end{figure}

@@

We have already suspected (see Table~\ref{table:simulation-aspects}) that in order to properly
estimate parameters of our system, we need to account for natural cell-variability, thus assigning
parameters on a per-cell basis.
We have previously presented a list of $N_p$ parameters (see Table~\ref{table:list-of-parameters})
which will either be assigned individually or globally.
We filter out a subset $N_g\leq N_p$ of global parameters, treating the remaining ones individually.
Given $N_c$ cells, the total amount of parameters to be estimated is:
\begin{equation}
    N_{p,\text{tot}} = N_g + N_c (N_p - N_g).
\end{equation}
It is clear, that we should avoid, to needlessly treat parameters non-globally since this increases
the size of the parameter space considerably.
The volume that needs to be sampled to optimize within a given parameter space scales exponentially
with the dimension ie. the total number of parameters $N_{p,\text{tot}}$.
\textbf{[Das ist leider komplett unverständlich. Ich glaube, ich weiss schon, was du sagen möchtest, aber das muss umformuliert werden. Stichworte sind hier: Cell Variability, Heterogeneity, latent variables. Parameter per cell and individual parameters versteht auch niemand den Unterschied.]}\\
\textit{Ich habe versucht, es klarer zu machen. Bitte dazu Feedback!}\\
\textbf{Es muss noch im Methoden Teil mehr erwähnt werden.}

@@

\textbf{Explain fit parameter. Explain $t_0$. Give citation for exponential growth. Also link to
equation in cell-cycle section}

@@

% --------------------------------------------------------------------------------------------------
\subsubsection{Without Division}

\begin{itemize}
    \item refer to which data?
    \item refer to table of all parameters which are fitted
    \item refer to equations which are fitted
    \item "fit full mechanical model"
    \item link to everything
\end{itemize}

In this section we will focus on a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant direct comparison scheme (see
\ref{paragraph:direct-comparison}) for our cost function.
We have chosen said interval such that effects related to the lag phase can be neglected which
allows us to sample a lower amount of paremeters thus improving runtime performance of our
optimization scheme.
